Fast Automatic Bayesian Cubature Using Lattice Sampling

### 2019-07-20 BayesianCubaturePaper2018 - Rev1b.tex
revising

\begin{theorem} \label{thm:fastparam}
Under assumptions \eqref{fastcompAssump}, the parameters and credible interval widths in Theorem \ref{thm:param} may be expressed in terms of the fast transforms of the function data, the first column of the Gram matrix, and $\vc_\vtheta$ as follows:
\begin{align}
\nonumber
m_\MLE &=  m_{\full} = m_{\GCV} =  \frac{\widetilde{y}_1}{n} = \frac 1n \sum_{i=1}^n y_i,
\\
\nonumber
s^2_\MLE 
& =
\frac{1}{n^2} 
\sum_{i=2}^n \frac{\abs{\widetilde{y}_i}^2}{\lambda_i}, \\
\nonumber
\widehat{\sigma}^2_{\textup{full}}
& =
\frac{1}{n(n-1)} \sum_{i=2}^n \frac{\abs{\widetilde{y}_i}^2}{\lambda_i}
\\
\nonumber
& \qquad \times
\left[\frac{\lambda_1}{n}{\left(1 - \frac{\widetilde{c}_1}{\lambda_1}\right)^2} + \left(c_0  - \frac 1n \sum_{i=1}^n \frac{\abs{\widetilde{c}_i}^2}{\lambda_i}\right) \right], \\
\nonumber 
s^2_{\textup{GCV}} & : =  \frac 1{n} \sum_{i=2}^n \frac{\abs{\widetilde{y}_i}^2}{\lambda_i^2}  \left [ \sum_{i=1}^n \frac{1}{\lambda_i} \right]^{-1},\\
\label{eqn_MLE_loss_func_optimized_2} %\nonumber
\vtheta_\MLE
&= 
\argmin_{\vtheta}
\left[
\log\left(
\sum_{i=2}^n \frac{\abs{\widetilde{y}_i}^2}{\lambda_{\vtheta i}}
\right) %\right .\\ \hspace{3cm} \left .
 + \frac{1}{n}\sum_{i=1}^n \log(\lambda_{\vtheta i})
\right],\\
\nonumber 
\vtheta_{\GCV} 
&= \argmin_\vtheta \left[ \log \left ( \sum_{i=2}^n \frac{\abs{\widetilde{y}_i}^2}{\lambda_{\vtheta i}^2} 
\right) \right .\\
\label{thetaGCV}
& \qquad \qquad \qquad \qquad  \left . -2\log\left( \sum_{i=1}^n \frac{1}{\lambda_{\vtheta i}} \right)
\right], \\
\nonumber
\hmu_\MLE  &= \hmu_{\full} = \hmu_{\GCV} =
\frac{\widetilde{y}_1}{n} +
\frac 1n \sum_{i=2}^n \frac{ \widetilde{c}_i^* \widetilde{y}_i}{\lambda_i}, \\
\nonumber
\err_\MLE  &
=
\frac{2.58}{n}\sqrt{
	\sum_{i=2}^{n} \frac{\abs{\widetilde{y}_i}^2}{\lambda_i}  
	\,
	\left( c_0 - \frac 1n \sum_{i=1}^n \frac{\abs{\widetilde{c}_i}^2}{\lambda_i} \right) 
}, \\
\nonumber
\err_{\full} & = t_{n-1,0.995} \hsigma_{\textup{full}}, \\
\nonumber
\err_{\textup{GCV}} & =
\frac{2.58}{n}\left\{\sum_{i=2}^n \frac{\abs{\widetilde{y}_i}^2}{\lambda_i^2}  \left [ \frac 1n \sum_{i=1}^n \frac{1}{\lambda_i} \right]^{-1} \right.
\\ 
\nonumber
&\qquad \qquad \left . \times
\left( c_0 - \frac 1n \sum_{i=1}^n \frac{\abs{\widetilde{c}_i}^2}{\lambda_i} \right) 
\right\}^{1/2}.
\end{align}
Under the further assumption \eqref{addAssump}, it follows that 
\begin{align}
\nonumber
\hmu_\MLE  &= \hmu_{\full} = \hmu_{\GCV} =
\frac{\widetilde{y}_1}{n} = \frac 1n \sum_{i=1}^n y_i,
\end{align}
and so $\hmu$ is simply the sample mean.  Also, under assumption \eqref{addAssump}, the credible interval widths simplify to
\begin{align}
\label{eq:errMLEAllAsump}
\err_\MLE  &
=
\frac{2.58}{n}\sqrt{
	\sum_{i=2}^{n} \frac{\abs{\widetilde{y}_i}^2}{\lambda_i}  
	\,
	\left( 1 -  \frac{n}{\lambda_1} \right) 
}, \\
\label{FJH:eq:errFullSimple}
\err_{\textup{full}}
&=
t_{n-1,0.995}
\sqrt{\frac{1}{n(n-1)} \sum_{i=2}^n \frac{\abs{\widetilde{y}_i}^2}{\lambda_i}  \left(\frac{\lambda_1}{n}  - 1  \right)}, \\
\nonumber
\err_{\textup{GCV}} & =
\frac{2.58}{n}\left\{\sum_{i=2}^n \frac{\abs{\widetilde{y}_i}^2}{\lambda_i^2}  \left [ \frac 1n \sum_{i=1}^n \frac{1}{\lambda_i} \right]^{-1} 
\right. \\ &\qquad \qquad \left . \times
\left( 1 -  \frac{n}{\lambda_1} \right)  
\right\}^{1/2}. \label{errGCVSimple}
\end{align}
\end{theorem}

@@

The set of nodes used is defined by a shifted extensible integration lattice node sequence, which takes the form
\begin{equation} \label{eqn:lattice_def}
\vx_{i} = \vh \phi(i-1) + \vDelta \mod \vone, \qquad i \in \naturals.
\end{equation} 
Here, $\vh$ is a $d$-dimensional generating vector of positive integers, $\vDelta$ is some point in $[0,1)^d$, often chosen at random, and $\{\phi(i)\}_{i=0}^n$ is the van der Corput sequence, defined by reflecting the binary digits of the integer about the decimal point, i.e., 
\begin{align} \label{vdCDef}
\begin{array}{r|ccccccccccccc}
i & 0 & 1 & 2 & 3 & 4 &  5 & 6 & 7 & \cdots \\
i & 0_2 & 1_2 & 10_2 & 11_2 & 100_2 & 101_2 & 110_2 & 111_2  & \cdots\\
\toprule
\phi(i) & {}_2.0 &  {}_2.1 & {}_2.01 &  {}_2.11  & {}_2.001 &  {}_2.101 & {}_2.011 &  {}_2.111 & \cdots\\
\phi(i) & 0 &  0.5 &  0.25 & 0.75 &  0.125 & 0.625  &  0.375 & 0.875 & \cdots
\end{array}
\end{align}
Note that 
\begin{multline} \label{phiprop}
\phi:\{0, \ldots, n-1 \} \to \{0, 1/n, \ldots, (n-1)/n\} \\
\text{is one-to-one},
\end{equation}
assuming $n$ is a power of $2$.

@@

Larger $r$ implies a greater degree of smoothness of the kernel.  Larger $\eta$ implies greater fluctuations of the output with respect to the input.  Plots of $C(\cdot, 0.3)$ are given in \figref{fig:fourierkernel-dim1} for various $r$ and $\eta$ values.

@@

Assumptions \eqref{fastcompAssumpA} and \eqref{fastcompAssumpB} can be verified by \eqref{PermMat}, \eqref{Keig}, and \eqref{Clateig}.  Assumption \eqref{fastcompAssumpC} is satisfied because $\mV^H \vb = \mW^H \mP^T \vb$ is just the permutation of the rows of the discrete Fourier transform of a vector.  This can be performed in $\Order(n \log(n))$ operations by the FFT algorithm. By a proper scaling of the kernel $K_\vtheta$ ensures that assumption \eqref{addAssump} is satisfied.

\subsection{Overcoming Cancellation Error}
For the kernels used in our computation, it may happen that $n/\lambda_1$ is close to $1$.  Thus, the term $1-n/\lambda_1$, which appears in the credible interval widths, $\err_{\MLE}$, $\err_{\textup{full}}$, and $\err_{\textup{GCV}}$, may suffer from cancellation error and even become negative.  We have observed this phenomenon.  We can avoid this cancellation error by modifying how we compute the Gram matrix and its eigenvalues.

@@

\begin{algorithm}
	\caption{Fast Automatic Bayesian Cubature}\label{algorithm2}
	\begin{algorithmic}[1]
		\Require a generator for the rank-1 Lattice sequence
		$\vx_1, \vx_2, \ldots$; 
		a shift-invariant periodic kernel, $C$;
		a black-box function, $f$; 
		an absolute error tolerance,
		$\varepsilon>0$; the positive initial sample size, $n_0$;
		the maximum sample size $n_{\textup{max}}$

@@

		\State\label{LoopStart}Generate $\{ \vx_i\}_{i=n' + 1}^{n}$ and sample $\{f(\vx_i)\}_{i=n'+1}^{n}$
		\State Compute $\vtheta$ by \eqref{thetaSimpleMLE} or \eqref{thetaSimpleGCV}
		\State Compute $\err_{\textup{CI}}$  according to \eqref{fastStoppingCriterionMLE}, \eqref{fastStoppingCriterionFull}, or \eqref{fastStoppingCriterionGCV}

@@

We summarize the results of this section and the previous one in the theorem below.  In comparison to Algorithm \ref{algorithm1}, the cost second and third components of the computational cost of Algorithm  \ref{algorithm2} are substantially reduced.
\begin{theorem}
Any symmetric, positive definite,  shift-\break invariant covariance kernel of the form \eqref{eq:shInv} scaled to satisfy \eqref{addAssump}, when matched with rank-1 lattice data-sites, must satisfy assumptions \eqref{fastcompAssump}.  The cubature, $\hmu$, is just the sample mean.  The \emph{fast Fourier transform} (FFT) can be used to expedite the estimates of $\vtheta$ in \eqref{thetaSimple} and the credible interval widths \eqref{fastStoppingCriterions} so that Algorithm \ref{algorithm2} has a computational cost which is the sum of the following:
\begin{itemize}
	\item $\Order\bigl(n\$(f) \bigr)$ for the function data, where $\$(f)$ is the computational cost of a single $f(\vx)$;

@@

	\item $\Order\bigl(N_{\opt} n \log(n) \bigr)$ for the FFT calculations.

@@

We have implemented the fast adaptive Bayesian cubature algorithm in MATLAB as part of the Guaranteed Adaptive Integration Library (GAIL) \cite{ChoEtal17b} as \allowbreak \code{cubBayesLattice\_g}. This algorithm uses the kernel defined in  \eqref{the_kernel_eqn_bernoulli} with  $r=1,2$ and the periodizing variable transforms in  \secref{period_var_tx}.  The rank-1 lattice node generator is taken from \cite{Nuy17a} (\code{exod2\_base2\_m20}).

@@

\subsection{Periodizing Variable Transformations}
\label{period_var_tx}
The shift-invariant covariance kernels underlying our Bayesian cubature  assume that the integrand has a degree of periodicity, with the smoothness assumed depending on the smoothness of the kernel.  While integrands arising in practice may be smooth, they might not be periodic.  Variable transformations can be used to ensure periodicity.

>>>>

\begin{theorem} \label{thm:fastparam}
Under assumptions \eqref{fastcompAssump}, the parameters and credible interval widths in Theorem \ref{thm:param} may be expressed in terms of the fast transforms of the function data, the first column of the Gram matrix, and $\vc_\vtheta$ as follows:
\begin{align}
\nonumber
m_\MLE &=  m_{\full} = m_{\GCV} =  \frac{\widetilde{y}_1}{n} = \frac 1n \sum_{i=1}^n y_i,
\\
\nonumber
s^2_\MLE 
& =
\frac{1}{n^2} 
\sum_{i=2}^n \frac{\abs{\widetilde{y}_i}^2}{\lambda_i}, \\
\nonumber
\widehat{\sigma}^2_{\textup{full}}
& =
\frac{1}{n(n-1)} \sum_{i=2}^n \frac{\abs{\widetilde{y}_i}^2}{\lambda_i}
\\
\nonumber
& \qquad \times
\left[\frac{\lambda_1}{n}{\left(1 - \frac{\widetilde{c}_1}{\lambda_1}\right)^2} + \left(c_0  - \frac 1n \sum_{i=1}^n \frac{\abs{\widetilde{c}_i}^2}{\lambda_i}\right) \right], \\
\nonumber 
s^2_{\textup{GCV}} & : =  \frac 1{n} \sum_{i=2}^n \frac{\abs{\widetilde{y}_i}^2}{\lambda_i^2}  \left [ \sum_{i=1}^n \frac{1}{\lambda_i} \right]^{-1},
\end{align}
\begin{subequations}
	\label{fastTheta}
	\begin{align}
\nonumber
\vtheta_\MLE
&= 
\argmin_{\vtheta}
\left[
\log\left(
\sum_{i=2}^n \frac{\abs{\widetilde{y}_i}^2}{\lambda_{\vtheta i}}
\right) \right .\\ 
\label{eqn_MLE_loss_func_optimized_2} %
& \hspace{3cm} \left .
 + \frac{1}{n}\sum_{i=1}^n \log(\lambda_{\vtheta i})
\right],\\
\nonumber 
\vtheta_{\GCV} 
&= \argmin_\vtheta \left[ \log \left ( \sum_{i=2}^n \frac{\abs{\widetilde{y}_i}^2}{\lambda_{\vtheta i}^2} 
\right) \right .\\
\label{thetaGCV}
& \qquad \qquad \qquad \qquad  \left . -2\log\left( \sum_{i=1}^n \frac{1}{\lambda_{\vtheta i}} \right)
\right], 
\end{align}
\end{subequations}
\begin{align}
\nonumber
\hmu_\MLE  &= \hmu_{\full} = \hmu_{\GCV} =
\frac{\widetilde{y}_1}{n} +
\frac 1n \sum_{i=2}^n \frac{ \widetilde{c}_i^* \widetilde{y}_i}{\lambda_i}, \\
\nonumber
\err_\MLE  &
=
\frac{2.58}{n}\sqrt{
	\sum_{i=2}^{n} \frac{\abs{\widetilde{y}_i}^2}{\lambda_i}  
	\,
	\left( c_0 - \frac 1n \sum_{i=1}^n \frac{\abs{\widetilde{c}_i}^2}{\lambda_i} \right) 
}, \\
\nonumber
\err_{\full} & = t_{n-1,0.995} \hsigma_{\textup{full}}, \\
\nonumber
\err_{\textup{GCV}} & =
\frac{2.58}{n}\left\{\sum_{i=2}^n \frac{\abs{\widetilde{y}_i}^2}{\lambda_i^2}  \left [ \frac 1n \sum_{i=1}^n \frac{1}{\lambda_i} \right]^{-1} \right.
\\ 
\nonumber
&\qquad \qquad \left . \times
\left( c_0 - \frac 1n \sum_{i=1}^n \frac{\abs{\widetilde{c}_i}^2}{\lambda_i} \right) 
\right\}^{1/2}.
\end{align}
Under the further assumption \eqref{addAssump}, it follows that 
\begin{equation}
\label{muhatGCV-FB-MLE-Simple}
\hmu_\MLE  = \hmu_{\full} = \hmu_{\GCV} =
\frac{\widetilde{y}_1}{n} = \frac 1n \sum_{i=1}^n y_i,
\end{equation}
and so $\hmu$ is simply the sample mean.  Also, under assumption \eqref{addAssump}, the credible interval widths simplify to
\begin{align}
\label{eq:errMLEAllAsump}
\err_\MLE  &
=
\frac{2.58}{n}\sqrt{
	\sum_{i=2}^{n} \frac{\abs{\widetilde{y}_i}^2}{\lambda_i}  
	\,
	\left( 1 -  \frac{n}{\lambda_1} \right) 
}, \\
\label{FJH:eq:errFullSimple}
\err_{\textup{full}}
&=
t_{n-1,0.995}
\sqrt{\frac{1}{n(n-1)} \sum_{i=2}^n \frac{\abs{\widetilde{y}_i}^2}{\lambda_i}  \left(\frac{\lambda_1}{n}  - 1  \right)}, \\
\nonumber
\err_{\textup{GCV}} & =
\frac{2.58}{n}\left\{\sum_{i=2}^n \frac{\abs{\widetilde{y}_i}^2}{\lambda_i^2}  \left [ \frac 1n \sum_{i=1}^n \frac{1}{\lambda_i} \right]^{-1} 
\right. \\ &\qquad \qquad \left . \times
\left( 1 -  \frac{n}{\lambda_1} \right)  
\right\}^{1/2}. \label{errGCVSimple}
\end{align}
\end{theorem}

@@

The set of nodes used is defined by a shifted extensible integration lattice node sequence, which takes the form
\begin{equation} \label{eqn:lattice_def}
\vx_{i} = \vh \phi(i-1) + \vDelta \mod \vone, \qquad i \in \naturals.
\end{equation} 
Here, $\vh$ is a $d$-dimensional generating vector of positive integers, $\vDelta$ is some point in $[0,1)^d$, often chosen at random, and $\{\phi(i)\}_{i=0}^n$ is the van der Corput sequence, defined by reflecting the binary digits of the integer about the decimal point, i.e., 
\begin{align} \label{vdCDef}
\begin{array}{r|ccccccccccccc}
i & 0 & 1 & 2 & 3 & 4 &  5 & 6 & 7 & \cdots \\
i & 0_2 & 1_2 & 10_2 & 11_2 & 100_2 & 101_2 & 110_2 & 111_2  & \cdots\\
\toprule
\phi(i) & {}_2.0 &  {}_2.1 & {}_2.01 &  {}_2.11  & {}_2.001 &  {}_2.101 & {}_2.011 &  {}_2.111 & \cdots\\
\phi(i) & 0 &  0.5 &  0.25 & 0.75 &  0.125 & 0.625  &  0.375 & 0.875 & \cdots
\end{array}
\end{align}
Note that 
\begin{multline} \label{phiprop}
\phi:\{0, \ldots, n-1 \} \to \{0, 1/n, \ldots, (n-1)/n\} \\
\text{is one-to-one},
\end{multline}
assuming $n$ is a power of $2$.

@@

Larger $r$ implies a greater degree of smoothness of the covariance kernel.  Larger $\eta$ implies greater fluctuations of the output with respect to the input.  Plots of $C(\cdot, 0.3)$ are given in \figref{fig:fourierkernel-dim1} for various $r$ and $\eta$ values.

@@

Assumptions \eqref{fastcompAssumpA} and \eqref{fastcompAssumpB} can be verified by \eqref{PermMat}, \eqref{Keig}, and \eqref{Clateig}.  Assumption \eqref{fastcompAssumpC} is satisfied because $\mV^H \vb = \mW^H \mP^T \vb$ is just the permutation of the rows of the discrete Fourier transform of a vector.  This can be performed in $\Order(n \log(n))$ operations by the FFT algorithm. A proper scaling of the kernel $K_\vtheta$ ensures that assumption \eqref{addAssump} is satisfied.

\subsection{Overcoming Cancellation Error}
For the covariance kernels used in our computation, it may happen that $n/\lambda_1$ is close to $1$.  Thus, the term $1-n/\lambda_1$, which appears in the credible interval widths, $\err_{\MLE}$, $\err_{\textup{full}}$, and $\err_{\textup{GCV}}$, may suffer from cancellation error and even become negative.  We have observed this phenomenon.  We can avoid this cancellation error by modifying how we compute the Gram matrix and its eigenvalues.

@@

\begin{algorithm}
	\caption{Fast Automatic Bayesian Cubature}\label{algorithm2}
	\begin{algorithmic}[1]
		\Require a generator for the rank-1 Lattice sequence
		$\vx_1, \vx_2, \ldots$; 
		a shift-invariant periodic kernel, $C_\vtheta$;
		a black-box function, $f$; 
		an absolute error tolerance,
		$\varepsilon>0$; the positive initial sample size, $n_0$;
		the maximum sample size $n_{\textup{max}}$

@@

		\State\label{LoopStart}Generate $\{ \vx_i\}_{i=n' + 1}^{n}$ and sample $\{f(\vx_i)\}_{i=n'+1}^{n}$
		\State Compute $\vtheta$ by \eqref{eqn_MLE_loss_func_optimized_2} or \eqref{thetaGCV}
		\State Compute $\err_{\textup{CI}}$  according to \eqref{fastStoppingCriterionMLE}, \eqref{fastStoppingCriterionFull}, or \eqref{fastStoppingCriterionGCV}

@@

We summarize the results of this section and the previous one in the theorem below.  In comparison to Algorithm \ref{algorithm1}, the cost second and third components of the computational cost of Algorithm  \ref{algorithm2} are substantially reduced.
\begin{theorem}
Any symmetric, positive definite,  shift-\break invariant covariance kernel of the form \eqref{eq:shInv} scaled to satisfy \eqref{addAssump}, when matched with rank-1 lattice data-sites, must satisfy assumptions \eqref{fastcompAssump}.  The cubature, $\hmu$, is just the sample mean.  The \emph{fast Fourier transform} (FFT) can be used to expedite the estimates of $\vtheta$ in \eqref{fastTheta} and the credible interval widths \eqref{fastStoppingCriterions} so that Algorithm \ref{algorithm2} has a computational cost which is the sum of the following:
\begin{itemize}
	\item $\Order\bigl(n\$(f) \bigr)$ for the function data, where $\$(f)$ is the computational cost of a single $f(\vx)$;

@@

	\item $\Order\bigl(N_{\opt} n \log(n) \bigr)$ for the FFT calculations; there is no $d$ dependence in these calculations.

@@

Although the third part of the computational cost has the largest dependence on $n$, in practice it need not be the largest contributor to the computational cost.  If function values are the result of an expensive simulation, then the first part may consume most of the computation time.

We have implemented the fast adaptive Bayesian cubature algorithm in MATLAB as part of the Guaranteed Adaptive Integration Library (GAIL) \cite{ChoEtal17b} as \allowbreak \code{cubBayesLattice\_g}. This algorithm uses the covariance kernel defined in  \eqref{the_kernel_eqn_bernoulli} with  $r=1,2$ and the periodizing variable transforms in  \secref{period_var_tx}.  The rank-1 lattice node generator is taken from \cite{Nuy17a} (\code{exod2\_base2\_m20}).

@@

\subsection{Periodizing Variable Transformations}
\label{period_var_tx}
The shift-invariant covariance kernels underlying our Bayesian cubature  assume that the integrand has a degree of periodicity, with the smoothness assumed depending on the smoothness of the covariance kernel.  While integrands arising in practice may be smooth, they might not be periodic.  Variable transformations can be used to ensure periodicity.
